Naturalness and Long-lived Beyond the Standard Model Particles

### 2017-07-21 tex/draft.tex
Rewrote naturalness, minor edits elsewhere

\usepackage{amsmath,amsthm,amssymb,mathtools,graphicx,caption,subcaption,mathrsfs,cancel,multirow,enumitem,setspace,sectsty,cite,hyperref}
\usepackage{fullpage}

@@

\section{The Standard Model}
    The standard model of particle physics (SM) describes all known particles and their non-gravitational interactions and is experimentally verified to astonishing (*weak word) accuracy (*say 2nd clause better). The SM consists of 12 spin-1/2 particles, the fermions, that make up all observed matter, 12 spin-1 particles, the gauge bosons, that communicate the electromagnetic, weak, and strong forces (*maybe put in pre-EWSB terms instead?), and 1 fundamental scalar, the Higgs boson, which will be described in detail below.

    All SM particles transform according to the gauged symmetry group SU(3)cxSU(2)LxU(1)Y, where SU(3)c is associated with color charge, SU(2)L is associated with weak isospin, and U(1)Y is associated with weak hypercharge. The gauge bosons are associated with the generators of these groups, thus there are 8 gluons associated with SU(3)c, 3 W bosons associated with SU(2)L, and 1 B boson associated with U(1)Y. The fermions are catagorized by which representation of each group they fit into. For example, SU(3)c distinguishes the quarks (which live in the fundamental representation) from the leptons (which are singlets), and SU(2)L distinguishes the left-handed fermions (doublets) from the right-handed fermions (singlets). Finally, different values of weak hypercharge further distignuish SM particles according to a linear combination of their weak isospin and electric charge (*is there some more fundamental way to describe Y?).

    The Higgs potential has a nonzero vacuum expectation value (VEV), which spontaneously breaks SU(2)LxU(1)Y to U(1)EM while leaving SU(3)c unscathed. Once the dust settles, the B and 3 W bosons have been mixed to produce the W+, W-, Z bosons that mediate the weak force and the photon that mediates electromagnetism. 3 of the 4 generators are broken in the process, leaving 3 extra degrees of freedom that ultimately become the masses of the W+, W-, and Z bosons while the remaining symmetry is associated with electric charge. Furthermore, expanding the Higgs field about its VEV produces mass terms for the quarks and charged leptons, thus giving mass to all massive SM particles. (*probably show a little math)

    (*kinda funny wording throughout. Maybe it would be better to trade some words for math?)
    All SM particles interact with virtual particles through loop interactions. These interactions lead to corrections to SM parameters such as particle masses. It is common to write a given parameter as a linear combination of its 'bare' (uncorrected) value and its corrections. As the only fundamental scalar in the SM, the Higgs is uniquely affected by these corrections. The next section explores the concept of naturalness, a major motivating principle for beyond the standard model (BSM) physics that is directly tied to the relationship between the Higgs bare mass and the terms that correct it.

\section{Naturalness and the Higgs mass}
    The naturalness criterion takes several forms. Here are a few (*cite these): (1) all parameters in a model should be of order 1, (2) Loop-level contributions should not be larger than tree-level contributions, (3) any small parameter should be protected by a custodial symmetry, and (4) fine-tuning is bad (*probably a better way to say 4). Although these statements are all motivated by the samefundamental notion, I find (3) and (4) to be most compelling.

    (*cite guidice naturally and t'hooft naturalness)
    Let's look at (3). In the SM, the small parameter of interest is the Higgs mass, which has no custodial symmetry to hold it down. This situation contrasts that of the fermions and gauge bosons, whose masses are kept small by custodial symmetries. That is, the SM would be more symmetric (*weird wording?) if fermions and gauge bosons were massless. Specifically, massless fermions would have an extra U(1)xU(1) chiral symmetry, and massless vector bosons would have an extra gauge symmetry (*?!?!?!). The Higgs, a fundamental scalar, has no such symmetry (*repetative) so we deem its low mass to be unnatural. (*explain why fundamental scalars have no custodial symmetry? I certainly better learn why.)

    Now look at (4), again in the context of the Higgs mass. The terms correcting the Higgs mass depend quadratically on the cutoff energy scale, which must be at or below the scale where quantum gravity becomes important (MP = 1/rootG ~ 10\^18??). This creates a sticky situation. If the terms correcting the bare mass are order(10\^32)GeV and the observed Higgs mass is order(100)GeV, then the bare mass must exactly cancel the corrections to approximately 1 part in 10\^30. At present, no good evidence exists to explain why the bare mass and corrections would match so perfectly. The level of exact cancellation is known as 'fine tuning'.

    (*probably doesn't need to be in final paper, but I'd like to check my basic understanding with Jamie, et al)
    There are other unnatural quantities that people worry about in physics. I know of two off the top of my head: (1) thetaQCD in the strong CP problem (2) the cosmological consant. The strong CP problem comes from the fact that QCD technically allows CP violation even though such violation has never been observed. The degree of violation is parameterized as thetaQCD, which recasts the problem into the naturalness context of 'why is thetaQCD so small?'. The cosmological constant is experimentally ~60 to ~120 orders of magnitude smaller than the predicted value (*predicted how?). While these issues deserve serious attention, they will not be considered further here.

    (*talk about anthropic principle or other arguments against?)
    (*I got tired of reading this section and haven't edited this paragraph at all yet)
    There are a few proposed solutions to the SM naturalness problem. Supersymmetry is the main one, and I'll spend the great majority of my time on it. Extra dimensions and composite Higgs models are the other two that pop up fairly often, but won't spend much time on them outside the next two sentences. Extra dimensions solves the hierarchy problem either by lowering the cutoff scale (as in large extra dimensions (ADD)) or introducing a warping factor that reduces the level of unnaturalness (warped extra dimensions (RS)). Composite Higgs models propose that the Higgs stops behaving as a fundamental scalar at some energy level, which removes the quadratic dependence. (*Hidden valley, dark sector, little Higgs?)

\section{Supersymmetry}
    Supersymmetry (SUSY) introduces a new symmetry in which every SM field fits into a larger multiplet with an inherent symmetry between bosons and fermions. In it's simplest form, SUSY results in one new boson for every SM fermion and one new fermion for every SM boson (*and an extra Higgs field - can I gloss over that?). The increase in particle multiplicity calls for a new naming convention: the spin-0 superpartners to the SM fermions are referred to as sfermions (e.g. stop or smuon), and the spin-1/2 superpartners to the SM bosons are given the name of their SM counterpart with "ino" tacked on to the end (e.g. Higgsino or wino). Together the superpartners to the SM particles are known as sparticles.

    When calculating contributions from loop diagrams, one finds that fermion loops differ in sign from boson loops, which means that in SUSY every bosonic correction to the Higgs mass is cancelled by a fermionic correction and vice versa. In fact, exact supersymmetry predicts that the bare Higgs mass is uncorrected to all orders in perturbation theory. (*show example diagram with contribution?)

@@

\section{Long-lived Particles}
    The SUSY variants most relevant to this discussion are those that are motivated by naturalness and (*and/or?) produce a signal that could be missed by common SUSY search techniques. Specifically, we are interested in SUSY models that predict long-lived particles (LLPs). That is, particles whose lifetimes are such that they decay at least a detectable distance from the nominal interaction point. This category includes everything from particles that decay at (*give ballparky number (~100um?)) to particles that propagate through the entire detector.

    In the SM and in BSM theories, LLPs result from one of the following conditions: (1) a symmetry that guarantees stability, (2) weak couplings to decay products, (3) highly-virtual intermediate states, or (4) limited phase space in which to decay. In the SM, charge and baryon number conservation guarantee the absolute stability of the electron and proton as the the lightest particles carrying their respective conserved quantities. Neutrons and muons, on the other hand, are long-lived beause the proton and electron masses leave them little phase space in which to decay (*also mention that they're both weak decays?). (*Mention neutrino, pion, kaon, some other bound state I might be forgetting?)

    In SUSY models, LLPs arise for the same basic reasons, and different SUSY models meet different LLP criteria in different ways. The most common (*based on my assumptions) SUSY LLP occurs in models with conserved R-partity, which is a proposed symmetry that protects many models from proton decay. In models with R-parity, the SM particles are assigned R-parity +1 and their superpartners are assigned R-parity -1. Conserving the product of R-parity at each vertex has two immediate phenomenalogical consequences: (1) sparticles can only be produced in pairs and (2) the lightest supersymmetric particle (LSP) is absolutely stable. A neutral LSP could show up in collider searches as significant missing energy, making MET the conventional SUSY signature in many searches.

    On the other hand, models with R-parity violation (RPV) produce long-lived, but not perfectly stable LSPs when RPV terms have small coupling constants. Luckily, RPV models can still avoid proton decay by allowing either terms that violate baryon number or terms that violate lepton number, but not both. A similar situation arises in GMSB models where the gravitino is the LSP. The coupling between the next-to-LSP (NLSP) and gravitino is inversely proportional to the scale of SUSY breaking, which supresses the NSLP decay rate and can lead to LLPs. (*is now a good time to explain GMSB?)(*cite dimopoulos low energy)

    LLPs from highly-virtual intermediate particles can occur in models such as split SUSY, where the scalar sparticles are significantly more massive than the gauginos. In these models, the gluino becomes long-lived when its decay to a quark and neutralino is mediated by a highly-virtual squark. (*cite kilian) (*mention that split throws out naturalness motivation?)

    The degenerate mass spectra of other SUSY models can also give rise to LLPs. In some AMSB models (*explain AMSB?), for example, the zino and wino are the LSP and NLSP, respectively, and are nearly degenerate in mass. The wino then becomes long-lived when its decay is suppressed by the lack of available phase space. (*cite R\&S out of this world)
    (*merge last 2 paragraphs under 'mass spectra lead to LLP' concept?)
    (*never said anything about hidden valley or dark sector stuff. Never said the word R-hadron. Never said stealth or displaced SUSY. Should I?)

\section{The CMS Detector}
\subsection{Overview}
    (*cite cms experiment at the cern lhc)
    The Compact Muon Solenoid (CMS) detector is designed to reconstruct a variety of particles from 14 TeV proton-proton collisions at the Large Hadron Collider (LHC). To meet this goal, CMS combines information from several subdetectors nested radially in and around a 4-T superconducting solenoid. All together, CMS is 21.6 m long, 14.6 m across, and weighs approximately 12500 tons. The subdetectors that make up CMS are described below.

\subsection{Tracker}
    (*cite CMS TDR, jinst 12 c02033)
    (*maybe I should trade some number listing for talking about how silicon detectors work)
    (*what about readout, cooling, etc?)
    In the region closest to the interaction point, CMS uses a fast, high-granularity silicon tracker to reconstruct tracks and pinpoint primary and secondary vertices. The tracker is subdivided into two regions: at r < 20 cm, the high particle flux (*specify) requires small pixels to keep occupancy near 0.1\%. Moving farther from the IP decreases particle flux, which allows the use of larger silicon strips.

    The pixel tracker, which was upgraded in winter 2016/17, makes up the innermost layer and features approximately 124 million silicon pixels, each of which covers an active area of 100 by 150 um\^2. The pixels are spread between four barrel layers (BPIX) at 2.9 < r < 16 cm and 3 endcap rings (FPIX) at 29 < z < 52 cm. The upgraded pixel tracker should provide 4-hit tracking out to eta < 2.4 and improved b-tagging over the original pixel tracker despite the increased pileup expected in 2017.

    Just outside the pixel tracker, the silicon strip tracker (SST) occupies the 20 < r < 116 cm and |z| < 282 cm region of CMS. Composed of 9.3 million silicon strips that vary from \_ to \_ in size (*look up) depending on particle flux in the region, the SST covers an active area of roughly 196 m\^2. At each trigger, the strip output voltage is transferred via optical link to be read out by off-detector front-end electronics (*right jargon?).

\subsection{Electromagnetic Calorimeter}
    (*cite CMS TDR)
    (*talk about readout?)
    (*More in the way of concept/how stuff works?)
    After traversing the inner tracker, a particle would next encounter the electromagnetic calorimeter (ECAL). As a homogeneous scintillation calorimeter, ECAL uses 61200 lead tunstate crystals in the barrel and 7324 in each endcap to reconstruct the energy deposited during electromagnetic showers. Lead tungstate crystals allow for a fast (80\% of light within 25 ns), compact (radiation length = 0.89 cm), fine-grained (Moliere length = 2.2 cm), and radiation hard (up to 10 Mrad(*learn context for this number)) calorimeter. The main drawback is the relatively low light yield (30 photon / MeV), which necesitates photodetectors with intrinsic gain that work in magnetic fields (*say B-field part better).

    The barrel section (EB) extends radially from 129 to 177 (*look up) cm and covers a pseudorapidity range of 0 to 1.479. The crystals are tapered to approximately project back to the IP but not so perfectly that likely particle trajectories align with cracks. Each crystal is approximately one Moliere radius wide and 25.8 radiation lengths deep.

    The crystals in each endcap section (EE) are arranged in an x-y grid that starts at |z| = 315 cm and covers a pseudorapidity range of 1.479 to 3.0. Additionally, a preshower detector, which consists of two planes of silicon strip detectors and a lead absorber, sits just inside the grid of lead tungstate crystals. (*Say what the preshower is for? Is it for pi0 rejection? If so, how and why?)

\subsection{Hadronic Calorimeter}
    (*cite CMS TDR)
    (*talk about readout?)
    The hadronic calorimeter (HCAL) uses 3.7mm plates of plastic scintillator interspersed within brass absorber to reconstruct the energy deposited during hadronic showers. Embedded wavelength-shifting fibers capture the scintillation light and transfer it to clear fibers so it can eventually be read out by hybrid photodiodes.

    The barrel section (HB) spans a pseudorapidty range of -1.4 to 1.4 and has 2304 towers (*check), each of which contains 14 (*check) brass plates and covers an area of 0.087 x 0.087 in eta-phi (*get a feel for this area). In addition, an extra layer (or two at eta = 0) of scintillator tiles sits just outside the solenoid. This extra layer, known as hadron outer (HO), spans a pseudorapidty range of -1.26 to 1.26 and increases the minimum effective HCAL interaction length to greater than 11.8.

    Each endcap spans a pseudorapidity range of 1.3 to 3.0 with 14 towers in eta and 5 to 10 degree phi segmentation. Also, a steel and quartz fiber forward calorimeter (HF) sits 11.2 m from the interaction pipe in a pseudorapidty range of 3.0 to 5.0. In HF, particles produce Cherenkov light when traversing the quartz fibers that run parallel to the beamline. (*say why HF is useful or mention crazy flux?)

\subsection{Solenoid}
    (*cite cms experiment at the cern lhc)
    CMS employs a 4 T superconducting solenoid to aid in the measurement of charged particle momenta. The solenoid consists of 2168 turns of Nb-Ti superconducting cable and has an inner diameter of 6 m and a length of 12.5 m. The flux returns through an iron yoke that also houses the muon system. The significant (*quantify?) bending power helps CMS meet its physics goals of unambiguous sign determination and momentum resolution of 10\% for muons up to 1 TeV.

\subsection{Muon System}
    (*cite CMS TDR)
    (*Specifically mention CMS design specs to meet physics goals?)
    The muon system (MS), which is the outermost subdetector, detects muons with 3 varieties of gaseous ionization detectors: drift tubes (DTs) in the barrel, cathode strip chambers (CSCs) in the endcaps, and resistive plate chambers (RPCs) in both the barrel and endcaps. The choice of detector varies between barrel and endcap due to the difference in neutron-induced background, muon rates, and magnetic field strength. The RPCs are useful in both the endcaps and barrel because their fast response and good time resolution allow unambiguous bunch crossing identification. Combining MS and tracker measurements improves the muon pT resolution at high momenta where the tracker pT resolution declines (*quantify or show std plot?).

    In the barrel detector (BD), four layers of DTs and and RPCs are embedded in the solenoid return yoke at 4.0, 4.9, 5.9, and 7.9 m. (*Say how DT works?) The maximum drift length is 2 cm and the single point resolution is 200 um. When reconstructing a vector describing the muon momentum, the phi precision is better than 100 um in position and 1 mrad in direction.

    Each muon endcap detector (ME) is composed of 234 CSCs and 4 RPC stations (*figure out exactly what RPC station means and say that instead). Each CSC has six gas gaps with radial cathode strips and nearly perpendicualar planes of anode wires. The wire signal from the ionization-induced electron avalanch is fast enough to use in the L1 trigger, while the slower (*quantify) cathode strip signal provides precise (*quantify) position measurements. All in all, the CSC spacial resolution is approximately 200 um and the angular resolution is approximately 10 mrad.

\subsection{Trigger}
    (*cite CMS TDR, jinst 12 c03021)
    (*Is entire detector read out at L1?)
    (*Mention fpgas in L1 or more details about HLT?)
    (*Something about particle flow?)
    (*Sloppy, but I'm getting tired)
    The trigger reduces the data writing rate from the 40 MHz collision rate to less than 1 kHz so that events can be written to tape. The rate reduction happens in two stages: Level-1 (L1) and High-Level Trigger (HLT). L1 uses input from ECAL, HCAL, and MS that represents physics objects (e/gamma, tau, jets, MET, and mu) and uses custom electronics to reduce the rate to ~100 kHz in 3.8 us. The HLT then uses a dedicated processor farm to further reduce the rate to the desired < 1 kHz.

\section{LHC Searches for Long-lived Particles}
\subsection{Overview}
    LLPs leave distinct signatures that differentiate LLP signals from SM backgrounds but also add complexity to LLP analyses. CMS and ATLAS were generally designed to detect and reconstruct stable particles from prompt decays, so LLP analyses often need to stray from the traditional analysis path in terms of object reconstruction (*really reconstruction? Maybe object definition?), background estimation, and which data tier to use (*can I say this more generally?). This paper will focus on four complementary LLP signatures: displaced vertices, disappearing tracks, slow-moving particles, and particles that stop completely before decaying.

>>>>

\usepackage[draft]{fixme}
\fxsetup{layout=footnote, marginclue}
\usepackage{amsmath,amsthm,amssymb,mathtools,graphicx,caption,subcaption,mathrsfs,cancel,multirow,enumitem,setspace,sectsty,cite,hyperref}
\usepackage{fullpage}

@@

\section{The Standard Model}
    The standard model of particle physics (SM) describes all known particles and their non-gravitational interactions and is experimentally verified to astonishing \fxnote{weak word} accuracy. The SM consists of 12 spin-1/2 particles, the fermions, that make up all observed matter, 12 spin-1 particles, the gauge bosons, that communicate the electromagnetic, weak, and strong forces, and 1 fundamental scalar, the Higgs boson, which will be described in detail below.

    All SM particles transform according to the gauged symmetry group SU(3)cxSU(2)LxU(1)Y, where SU(3)c is associated with color charge, SU(2)L is associated with weak isospin, and U(1)Y is associated with weak hypercharge. The gauge bosons are associated with the generators of these groups, thus there are 8 gluons associated with SU(3)c, 3 W bosons associated with SU(2)L, and 1 B boson associated with U(1)Y. The fermions are catagorized by which representation of each group they fit into. For example, SU(3)c distinguishes the quarks (which live in the fundamental representation) from the leptons (which are singlets), and SU(2)L distinguishes the left-handed fermions (doublets) from the right-handed fermions (singlets). Finally, different values of weak hypercharge further distignuish SM particles according to a linear combination of their weak isospin and electric charge \fxnote{is there some more fundamental way to describe Y?}.

    The Higgs potential has a nonzero vacuum expectation value (VEV), which spontaneously breaks SU(2)LxU(1)Y to U(1)EM while leaving SU(3)c unscathed. Once the dust settles, the B and 3 W bosons have been mixed to produce the W+, W-, Z bosons that mediate the weak force and the photon that mediates electromagnetism. 3 of the 4 generators are broken in the process, leaving 3 extra degrees of freedom that ultimately become the masses of the W+, W-, and Z bosons while the remaining symmetry is associated with electric charge. Furthermore, expanding the Higgs field about its VEV produces mass terms for the quarks and charged leptons, thus giving mass to all massive SM particles. \fxnote{probably show a little math}

    \fxnote{kinda funny wording throughout. Maybe it would be better to trade some words for math?}
    All SM particles interact with virtual particles through loop interactions. These interactions lead to corrections to SM parameters such as particle masses. It is common to write a given parameter as a linear combination of its 'bare' (uncorrected) value and its corrections. As the only fundamental scalar in the SM, the Higgs is uniquely affected by these corrections. The next section explores the concept of naturalness, a major motivating principle for beyond the standard model (BSM) physics that is directly tied to the relationship between the Higgs bare mass and the terms that correct it.

\section{Naturalness and the Higgs mass}
    \fxnote{cite guidice naturally and t'hooft naturalness}
    \fxnote{Haven't specifically said 'hierarchy problem. Is that bad?}
    \fxnote{Never really talked about the effective theory concept or dimensional analysis}
    The naturalness criterion can be stated in many ways, a few of which are reproduced here: (1) All dimensionless parameters should be of order 1, (2) An effective theory should not be sensitive to small variations in the parameters of higher-energy theory, and (3) any dimensionless parameter much smaller than 1 must be protected by a custodial symmetry. When applied to the Higgs mass, naturalness leads to what's known as the 'hierarchy problem' or unacceptable levels of 'fine tuning'. These issues are unpacked below.

    Experimentally, the mass of the Higgs boson is 125 GeV, a value that must be achieved by a linear combination of the bare mass and loop diagram corrections. The scale of the loop corrections is proportional to the square of the cutoff scale of the SM, which is the highest energy at which the SM is valid. If, hypothetically, the SM is valid all the way up to the Planck scale (where quantum gravity becomes important), then Higgs mass corrections would be order($MP^2$) ~ $10^{32}$ GeV.

    Order($10^{32}$) corrections imply that the bare Higgs mass must be negative and also order($10^{32}$) to reproduce the experimetnally known value. In fact, the bare mass and corrections must match to 1 part in ~$10^{30}$. While it's entirely possible that these two numbers just so happen to match so perfectly, but many \fxnote{weak} feel that such miraculous fine tuning must be a hint that some deeper physical mechanism is at work. That is, either some previously unknown symmetry is protecting the Higgs mass or the cutoff scale is order(TeV).

    Before digging into possible solutions to this problem, one other question must be addressed: why doesn't this same argument apply to all SM particles? The answer comes from naturalness statement (3) above. As a fundamental scalar, the Higgs boson lacks the chiral or gauge symmetries enjoyed by the fermions and vector bosons, respectively. These two symmetries cause the symmetry of the SM to be enhanced when the fermion and vector boson masses are set to 0. \fxnote{work out fermion case?} Setting the higgs mass to 0 in no way increases the symmetry of the theory, so we have no reason to expect it to be small.

    There are three main approaches to resolving this problem: (1) argue that it was never really a problem in the first place, (2) reduce the SM cutoff scale, and (3) introduce a new symmetry to protect the Higgs mass. Arguments against naturalness often involve the anthropic principle, which claims that it should come as no surprise that humans live in a universe whose fundamental parameters are tuned to produce carbon-based life. Alternatively, some theories, such as large extra dimensions, lower the SM cutoff by proposing that gravity is spread across more than 3 spatial dimensions, which lowers the Planck mass itself. Finally, the dominant BSM theory that aims to solve the naturalness problem by introducing a new symmetry is supersymmetry, which is explained below. \fxnote{didn't mention composite higgs}

    \fxnote{Optional paragraph about non-Higgs unnaturalness:    There are other unnatural quantities that people worry about in physics. I know of two off the top of my head: (1) thetaQCD in the strong CP problem (2) the cosmological consant. The strong CP problem comes from the fact that QCD technically allows CP violation even though such violation has never been observed. The degree of violation is parameterized as thetaQCD, which recasts the problem into the naturalness context of 'why is thetaQCD so small?'. The cosmological constant is experimentally ~60 to ~120 orders of magnitude smaller than the predicted value \fxnote{predicted how?}. While these issues deserve serious attention, they will not be considered further here}

\section{Supersymmetry}
    Supersymmetry (SUSY) introduces a new symmetry in which every SM field fits into a larger multiplet with an inherent symmetry between bosons and fermions. In it's simplest form, SUSY results in one new boson for every SM fermion and one new fermion for every SM boson \fxnote{and an extra Higgs field - can I gloss over that?}. The increase in particle multiplicity calls for a new naming convention: the spin-0 superpartners to the SM fermions are referred to as sfermions (e.g. stop or smuon), and the spin-1/2 superpartners to the SM bosons are given the name of their SM counterpart with "ino" tacked on to the end (e.g. Higgsino or wino). Together the superpartners to the SM particles are known as sparticles.

    When calculating contributions from loop diagrams, one finds that fermion loops differ in sign from boson loops, which means that in SUSY every bosonic correction to the Higgs mass is cancelled by a fermionic correction and vice versa. In fact, exact supersymmetry predicts that the bare Higgs mass is uncorrected to all orders in perturbation theory. \fxnote{show example diagram with contribution?}

@@

\section{Long-lived Particles}
    The SUSY variants most relevant to this discussion are those that are motivated by naturalness and \fxnote{and/or?} produce a signal that could be missed by common SUSY search techniques. Specifically, we are interested in SUSY models that predict long-lived particles (LLPs). That is, particles whose lifetimes are such that they decay at least a detectable distance from the nominal interaction point. This category includes everything from particles that decay at \fxnote{give ballparky number (~100um?)} to particles that propagate through the entire detector.

    In the SM and in BSM theories, LLPs result from one of the following conditions: (1) a symmetry that guarantees stability, (2) weak couplings to decay products, (3) highly-virtual intermediate states, or (4) limited phase space in which to decay. In the SM, charge and baryon number conservation guarantee the absolute stability of the electron and proton as the the lightest particles carrying their respective conserved quantities. Neutrons and muons, on the other hand, are long-lived beause the proton and electron masses leave them little phase space in which to decay \fxnote{also mention that they're both weak decays?}. \fxnote{Mention neutrino, pion, kaon, some other bound state I might be forgetting?}

    In SUSY models, LLPs arise for the same basic reasons, and different SUSY models meet different LLP criteria in different ways. The most common \fxnote{based on my assumptions} SUSY LLP occurs in models with conserved R-partity, which is a proposed symmetry that protects many models from proton decay. In models with R-parity, the SM particles are assigned R-parity +1 and their superpartners are assigned R-parity -1. Conserving the product of R-parity at each vertex has two immediate phenomenalogical consequences: (1) sparticles can only be produced in pairs and (2) the lightest supersymmetric particle (LSP) is absolutely stable. A neutral LSP could show up in collider searches as significant missing energy, making MET the conventional SUSY signature in many searches.

    On the other hand, models with R-parity violation (RPV) produce long-lived, but not perfectly stable LSPs when RPV terms have small coupling constants. Luckily, RPV models can still avoid proton decay by allowing either terms that violate baryon number or terms that violate lepton number, but not both. A similar situation arises in GMSB models where the gravitino is the LSP. The coupling between the next-to-LSP (NLSP) and gravitino is inversely proportional to the scale of SUSY breaking, which supresses the NSLP decay rate and can lead to LLPs. \fxnote{is now a good time to explain GMSB?}\fxnote{cite dimopoulos low energy}

    LLPs from highly-virtual intermediate particles can occur in models such as split SUSY, where the scalar sparticles are significantly more massive than the gauginos. In these models, the gluino becomes long-lived when its decay to a quark and neutralino is mediated by a highly-virtual squark. \fxnote{cite kilian} \fxnote{mention that split throws out naturalness motivation?}

    The degenerate mass spectra of other SUSY models can also give rise to LLPs. In some AMSB models \fxnote{explain AMSB?}, for example, the zino and wino are the LSP and NLSP, respectively, and are nearly degenerate in mass. The wino then becomes long-lived when its decay is suppressed by the lack of available phase space. \fxnote{cite R\&S out of this world}
    \fxnote{merge last 2 paragraphs under 'mass spectra lead to LLP' concept?}
    \fxnote{never said anything about hidden valley or dark sector stuff. Never said the word R-hadron. Never said stealth or displaced SUSY. Should I?}

\section{The CMS Detector}
\subsection{Overview}
    \fxnote{cite cms experiment at the cern lhc}
    The Compact Muon Solenoid (CMS) detector is designed to reconstruct a variety of particles from 14 TeV proton-proton collisions at the Large Hadron Collider (LHC). To meet this goal, CMS combines information from several subdetectors nested radially in and around a 4-T superconducting solenoid. All together, CMS is 21.6 m long, 14.6 m across, and weighs approximately 12500 tons. The subdetectors that make up CMS are described below.

\subsection{Tracker}
    \fxnote{cite CMS TDR, jinst 12 c02033}
    \fxnote{maybe I should trade some number listing for talking about how silicon detectors work}
    \fxnote{what about readout, cooling, etc?}
    In the region closest to the interaction point, CMS uses a fast, high-granularity silicon tracker to reconstruct tracks and pinpoint primary and secondary vertices. The tracker is subdivided into two regions: at r < 20 cm, the high particle flux \fxnote{specify} requires small pixels to keep occupancy near 0.1\%. Moving farther from the IP decreases particle flux, which allows the use of larger silicon strips.

    The pixel tracker, which was upgraded in winter 2016/17, makes up the innermost layer and features approximately 124 million silicon pixels, each of which covers an active area of 100 by 150 $um^2$. The pixels are spread between four barrel layers (BPIX) at 2.9 < r < 16 cm and 3 endcap rings (FPIX) at 29 < z < 52 cm. The upgraded pixel tracker should provide 4-hit tracking out to eta < 2.4 and improved b-tagging over the original pixel tracker despite the increased pileup expected in 2017.

    Just outside the pixel tracker, the silicon strip tracker (SST) occupies the 20 < r < 116 cm and |z| < 282 cm region of CMS. Composed of 9.3 million silicon strips that vary from \_ to \_ in size \fxnote{look up} depending on particle flux in the region, the SST covers an active area of roughly 196 $m^2$. At each trigger, the strip output voltage is transferred via optical link to be read out by off-detector front-end electronics \fxnote{right jargon?}.

\subsection{Electromagnetic Calorimeter}
    \fxnote{cite CMS TDR}
    \fxnote{talk about readout?}
    \fxnote{More in the way of concept/how stuff works?}
    After traversing the inner tracker, a particle would next encounter the electromagnetic calorimeter (ECAL). As a homogeneous scintillation calorimeter, ECAL uses 61200 lead tunstate crystals in the barrel and 7324 in each endcap to reconstruct the energy deposited during electromagnetic showers. Lead tungstate crystals allow for a fast (80\% of light within 25 ns), compact (radiation length = 0.89 cm), fine-grained (Moliere length = 2.2 cm), and radiation hard (up to 10 Mrad\fxnote{learn context for this number}) calorimeter. The main drawback is the relatively low light yield (30 photon / MeV), which necesitates photodetectors with intrinsic gain that work in magnetic fields \fxnote{say B-field part better}.

    The barrel section (EB) extends radially from 129 to 177 \fxnote{look up} cm and covers a pseudorapidity range of 0 to 1.479. The crystals are tapered to approximately project back to the IP but not so perfectly that likely particle trajectories align with cracks. Each crystal is approximately one Moliere radius wide and 25.8 radiation lengths deep.

    The crystals in each endcap section (EE) are arranged in an x-y grid that starts at |z| = 315 cm and covers a pseudorapidity range of 1.479 to 3.0. Additionally, a preshower detector, which consists of two planes of silicon strip detectors and a lead absorber, sits just inside the grid of lead tungstate crystals. \fxnote{Say what the preshower is for? Is it for pi0 rejection? If so, how and why?}

\subsection{Hadronic Calorimeter}
    \fxnote{cite CMS TDR}
    \fxnote{talk about readout?}
    The hadronic calorimeter (HCAL) uses 3.7mm plates of plastic scintillator interspersed within brass absorber to reconstruct the energy deposited during hadronic showers. Embedded wavelength-shifting fibers capture the scintillation light and transfer it to clear fibers so it can eventually be read out by hybrid photodiodes.

    The barrel section (HB) spans a pseudorapidty range of -1.4 to 1.4 and has 2304 towers \fxnote{check}, each of which contains 14 \fxnote{check} brass plates and covers an area of 0.087 x 0.087 in eta-phi \fxnote{get a feel for this area}. In addition, an extra layer (or two at eta = 0) of scintillator tiles sits just outside the solenoid. This extra layer, known as hadron outer (HO), spans a pseudorapidty range of -1.26 to 1.26 and increases the minimum effective HCAL interaction length to greater than 11.8.

    Each endcap spans a pseudorapidity range of 1.3 to 3.0 with 14 towers in eta and 5 to 10 degree phi segmentation. Also, a steel and quartz fiber forward calorimeter (HF) sits 11.2 m from the interaction pipe in a pseudorapidty range of 3.0 to 5.0. In HF, particles produce Cherenkov light when traversing the quartz fibers that run parallel to the beamline. \fxnote{say why HF is useful or mention crazy flux?}

\subsection{Solenoid}
    \fxnote{cite cms experiment at the cern lhc}
    CMS employs a 4 T superconducting solenoid to aid in the measurement of charged particle momenta. The solenoid consists of 2168 turns of Nb-Ti superconducting cable and has an inner diameter of 6 m and a length of 12.5 m. The flux returns through an iron yoke that also houses the muon system. The significant \fxnote{quantify?} bending power helps CMS meet its physics goals of unambiguous sign determination and momentum resolution of 10\% for muons up to 1 TeV.

\subsection{Muon System}
    \fxnote{cite CMS TDR}
    \fxnote{Specifically mention CMS design specs to meet physics goals?}
    The muon system (MS), which is the outermost subdetector, detects muons with 3 varieties of gaseous ionization detectors: drift tubes (DTs) in the barrel, cathode strip chambers (CSCs) in the endcaps, and resistive plate chambers (RPCs) in both the barrel and endcaps. The choice of detector varies between barrel and endcap due to the difference in neutron-induced background, muon rates, and magnetic field strength. The RPCs are useful in both the endcaps and barrel because their fast response and good time resolution allow unambiguous bunch crossing identification. Combining MS and tracker measurements improves the muon pT resolution at high momenta where the tracker pT resolution declines \fxnote{quantify or show std plot?}.

    In the barrel detector (BD), four layers of DTs and and RPCs are embedded in the solenoid return yoke at 4.0, 4.9, 5.9, and 7.9 m. \fxnote{Say how DT works?} The maximum drift length is 2 cm and the single point resolution is 200 um. When reconstructing a vector describing the muon momentum, the phi precision is better than 100 um in position and 1 mrad in direction.

    Each muon endcap detector (ME) is composed of 234 CSCs and 4 RPC stations \fxnote{figure out exactly what RPC station means and say that instead}. Each CSC has six gas gaps with radial cathode strips and nearly perpendicualar planes of anode wires. The wire signal from the ionization-induced electron avalanch is fast enough to use in the L1 trigger, while the slower \fxnote{quantify} cathode strip signal provides precise \fxnote{quantify} position measurements. All in all, the CSC spacial resolution is approximately 200 um and the angular resolution is approximately 10 mrad.

\subsection{Trigger}
    \fxnote{cite CMS TDR, jinst 12 c03021}
    \fxnote{Is entire detector read out at L1?}
    \fxnote{Mention fpgas in L1 or more details about HLT?}
    \fxnote{Something about particle flow?}
    \fxnote{Sloppy, but I'm getting tired}
    The trigger reduces the data writing rate from the 40 MHz collision rate to less than 1 kHz so that events can be written to tape. The rate reduction happens in two stages: Level-1 (L1) and High-Level Trigger (HLT). L1 uses input from ECAL, HCAL, and MS that represents physics objects (e/gamma, tau, jets, MET, and mu) and uses custom electronics to reduce the rate to ~100 kHz in 3.8 us. The HLT then uses a dedicated processor farm to further reduce the rate to the desired < 1 kHz.

\section{LHC Searches for Long-lived Particles}
\subsection{Overview}
    LLPs leave distinct signatures that differentiate LLP signals from SM backgrounds but also add complexity to LLP analyses. CMS and ATLAS were generally designed to detect and reconstruct stable particles from prompt decays, so LLP analyses often need to stray from the traditional analysis path in terms of object reconstruction \fxnote{really reconstruction? Maybe object definition?}, background estimation, and which data tier to use \fxnote{can I say this more generally?}. This paper will focus on four complementary LLP signatures: displaced vertices, disappearing tracks, slow-moving particles, and particles that stop completely before decaying.
